\silentchapter{Three Letters}{three letters}
\begin{widebody}
\begin{foreordbody}

\vspace{1mm}
\noindent{\scshape\addfontfeature{LetterSpace=6}\small to the designer}
\vspace{1mm}

You already know the diagnosis: Palantir, the disposable cup, the architecture of attention. What the book offers is not method, but vocabulary. With it you can document what you took into account, what you chose away, and why. No form-maker works without axes that are chosen away; the difference between the field as it is and the field as it could be lies in whether the omission is hidden or made visible. Begin with the next deliverable.

\vspace{3mm}
\noindent{\scshape\addfontfeature{LetterSpace=6}\small to the student}
\vspace{1mm}

The curriculum is under construction; what you demand of your studies shapes what the field becomes. Demand written justification for every design choice: which axes were represented, which were chosen away. Demand time in the substrate, not only in the representation. Demand consultation with neighbouring fields where the form touches living systems, large populations, ecology, power. A candidate who has never said no to a commission has not completed her education; she has completed a copying.

\vspace{3mm}
\noindent{\scshape\addfontfeature{LetterSpace=6}\small to the academy}
\vspace{1mm}

The framework is falsifiable; the research programme has four points. What is missing is not theory, but institutional commitment: an organ that can exclude for violation, a curriculum that binds the craft to substrate and neighbouring fields. Sustainability and climate are already required in higher education; these requirements lead directly to ecology, cybernetics, Bateson, and systems thinking, frameworks already integrated in this foundation. Here lies an immediate action point: bind existing sustainability requirements to the conceptual tools that make them operational. The archive is enormous. It is underutilised. It is waiting.

\vspace{3mm}

The care template on the next page is a tool the designer can consult. It lists the scales from quark to civilisation,\textsuperscript{85,\,87,\,91} each row an agent with problem space, time span, and care requirement. The row above carries the form; the row below carries it. The template is also a research-ethical tool for interdisciplinary collaboration.

\end{foreordbody}
\end{widebody}
