\section{The Theory in Action: Three Worked Examples}
\label{sec:examples}

The empirical demonstration in Section~\ref{sec:empirical} tests the theory's predictions against a historical corpus. But a theory of design form must also illuminate contemporary practice and domains beyond furniture. This section applies the propositions to three concrete cases: one historical, one contemporary-biological, and one continental. The examples are \textit{illustrative}, not probative; they demonstrate the theory's explanatory reach and generate testable predictions, but they do not constitute independent empirical tests. Each example is chosen to test a specific aspect of the theory: the first tests convergence (Propositions 1--4), the second tests multi-scale agency (Propositions 5--6), and the third tests the care diagnostic (Proposition 7).

\subsection{The cantilever chair race, 1925--1928}

Between 1925 and 1928, four designers working in three cities independently converged on a shared morphological innovation: the cantilevered tubular steel chair.

Marcel Breuer designed the Wassily (B3) in 1925 at the Bauhaus in Dessau, using cold-bent nickel-plated steel tubing inspired by his Adler bicycle. Breuer's cognitive cone spanned visual lightness, Bauhaus pedagogy, and production economy; he attended to how the chair \textit{looked} in the pedagogical space. Mart Stam presented the first true cantilever (S33) at a Stuttgart housing exhibition in 1926, using straight gas-pipe segments joined at right angles. Stam's cone was oriented toward structural minimalism and socialist housing policy; he attended to how the chair \textit{could be produced cheaply}. Ludwig Mies van der Rohe exhibited his cantilevered MR chair in 1927 at the Weissenhof exhibition, using a continuous curved tube that gave the seat its characteristic spring. Mies's cone encompassed material elegance and spatial continuity; he attended to how the chair \textit{behaved} in space. Breuer returned with the Cesca (B32) in 1928, combining a cantilevered steel frame with a caned seat and back.

The resulting forms differ precisely along the axes each designer attended to. Stam's S33 is rigid and austere; Mies's MR is elastic and sensual; Breuer's Cesca is hybrid and pragmatic. The convergence toward cantilever; and the divergence along care axes; is exactly what the theory predicts.

In the theory's terms, the material affordance of cold-bent steel tubing had been latent for over forty years (Proposition 2). Steel tubing was manufactured for bicycles and plumbing from the 1880s onward. The \textit{operation}; cold bending of thin-walled tubes to furniture scale; was the missing link. The convergence of four designers within three years confirms that the landscape attractor was real (Proposition 1): the fitness peak existed in the landscape before anyone climbed it. The lawsuits that followed between Stam, Breuer, and Mies over priority are themselves evidence that independent agents converged on the same landscape feature from different directions.

\subsection{Neri Oxman's Silk Pavilion, MIT Media Lab, 2013}

The Silk Pavilion is a dome-shaped structure produced through a two-phase process. In the first phase, a CNC-controlled machine deposited a scaffold of silk threads across a steel frame, following a computationally generated pattern. In the second phase, 6,500 silkworms were placed on the scaffold and deposited additional silk over 20 days, guided by environmental parameters including light intensity and temperature gradients.

The result is a structure whose final form was not drawn by the designer. Neri Oxman and her team at the MIT Media Lab Mediated Matter group set the boundary conditions: scaffold geometry, species selection, environmental parameters. The biological agents navigated the form grammar at cellular scale.

The competence hierarchy is literal, not metaphorical. Each silkworm is a zero-order agent in the theory's sense (Proposition 5): it possesses a goal state (spinning a cocoon), a distance measure (proximity to suitable attachment points), and an adjustment mechanism (silk extrusion direction). The worm does not represent the dome; it responds to the local gradient. The hierarchy spans five timescales:

\begin{enumerate}
\item \textit{Cellular silk production} (millisecond scale): protein synthesis and extrusion.
\item \textit{Individual worm navigation} (minute scale): each worm selects direction by local light gradients.
\item \textit{Collective pattern formation} (day scale): 6,500 worms produce emergent texture.
\item \textit{Scaffold specification} (month scale): the designer defines the computational frame structure.
\item \textit{Material and species selection} (year scale): the research group selects silkworm over spider, steel over wood.
\end{enumerate}

\noindent Form arises in the intersection of these nested cognitive cones. The designer's care determines which axes are represented at the macro scale; the silkworms' biology handles the rest. The resulting form shows how far care reached at \textit{every} scale: the scaffold's precision at the computational scale, the organic texture at the biological scale. The Silk Pavilion is not an example of design \textit{with} biological materials; it is an example of design \textit{between} agents at different competence scales (Proposition 6).

\subsection{The Great Green Wall: continental-scale ecological design}

The Great Green Wall initiative, launched by the African Union in 2007, aims to restore 100 million hectares of degraded land across an 8,000-kilometer belt from Senegal to Djibouti. The project addresses desertification, food insecurity, climate change, and forced migration across the Sahel. As of 2024, approximately 18\% of the target area has been restored, with significant variation in approach and success across the eleven participating nations.

This is design at a scale that breaks the conventions of the design field. The ``form'' under design is not a single object but a multi-scalar configuration: soil microbiome composition (cellular-scale agent), root architecture and mycorrhizal networks (organism-scale agent), canopy structure and species diversity (community-scale agent), pastoral land-use patterns and community governance (social-scale agent), continental precipitation dynamics and carbon cycling (system-scale agent).

The theory predicts (Proposition 4) that collapsing the landscape to a single selection pressure; say, carbon sequestration; will produce configurations that fail under pressures the design was blind to. History confirms this. Early phases of the initiative focused heavily on tree planting (a single-pressure approach), and many plantations failed because they did not account for local soil conditions, pastoral mobility patterns, or indigenous species selection. The most successful restoration sites; such as the farmer-managed natural regeneration programs in Niger; employed a multi-pressure approach that coordinated biological, social, and economic agents simultaneously.

The theory reads this as a care problem (Proposition 7). The dimensionality of the cognitive cone must match the dimensionality of the pressure field. A design that represents only the carbon axis is an act of narrow care. A design that represents carbon, biodiversity, water table dynamics, pastoral rights, indigenous knowledge, and intergenerational equity is an act of expansive care. The form of the restored landscape shows how far care reached; and the failures show where it did not.

Here lies the theory's strongest generality test. If the core equation $\mathrm{Form} = SG(\nabla_{C(A)} L(c,t))$ holds for continental ecology, it holds beyond furniture. The grammar here is not Stiny's shape algebra but the ecological repertoire of operations (species selection, soil improvement techniques, irrigation systems). The landscape is not aesthetic-ergonomic but biophysical-social. The cognitive cone is not a single designer but a hierarchy of agents from soil microbe to intercontinental policy. The structure is the same; the content is different. The theory's ambition is that this structure is invariant.
